---
title: "Applied Metrics Empirical P-Set 2: Fall 2023"
output: pdf_document
date: "2023-11-29"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

### Question 1 - Estimate the propensity score by a probit regression
```{r }
# Clear all
rm(list = ls())

# load and require packages
pacman::p_load(foreign, tidyverse, here, randomForest)

# Load data
data <- read.csv(here("data/anya_cleaned.csv"))

# Fit probit
probit_model <- glm(d ~ I(racem) + I(agem) + agemomfirst + I(sexk) + z, 
                    data = data, 
                    family = binomial(link = "probit"))

# Predict propensity scores
data$prop_score <- predict(probit_model, type = "response")

```


### Question 2 - Estimate by OLS

```{r }
# Fit OLS
ols <- lm(lnincome ~ I(racem) + I(agem) + agemomfirst + I(sexk) + 
            prop_score*(I(racem) + I(agem) + agemomfirst + I(sexk)) + 
            prop_score + (prop_score)^2 + prop_score^3, 
                    data = data)

summary(ols)
```

### Question 2a - Relate structural equation and probit regression

Suppose we specify the potential outcome equations as $Y_i(1) = X'_i \theta_1 + \epsilon_{1i}$ and $Y_i(0) = X'_i \theta_0 + \epsilon_{0i}$, and the structural selection equation as $D_i = 1 \{u(X_i,Z_i) \geq V_i \}$. where $u(X_i,Z_i) - V_i$ is the latent utility derived from additional children.

The structural selection equation describes the switch from $D = 0$ to $D=1$ in terms of utility, i.e., whether, as a function of your observables and instrument, the family has a higher utility than some level of having more than three children. The propensity score quantifies this probability, capturing the likelihood, given your observables and instrument value, of having $D = 1$. 



### Question 2b - Assumptions for MTE

Clarify a set of assumptions that the marginal treatment effect $M(x,u)$, $u \in (0,1)$ can be identified by the derivative of the OLS regression equation WRT the propensity score: 
$$MTE(x, u) = \frac{\partial}{\partial p}\bigg|_{p=u} \left( x'\beta_0 + px'\beta_1 + \kappa_1p + \kappa_2p^2 + \kappa_3p^3 \right)$$

The assumptions are: XXXXX




### Question 2c - Estimate propensity score

Estimate the propensity score as instructed above, and assess whether the coefficient of the instrumental variable in the probit regression is significantly different from zero or not.

Yes, we see that the z-value for the instrument is extremely large at 49, and very significant. So we have confidence that the instrument is predicting treatment. 

```{r }
summary(probit_model)

```



### Question 2d - Plot histogram of estimated propensity scores

```{r }
hist(data$prop_score, 
     main = "Histogram of Estimated Propensity Scores", 
     xlab = "Propensity Score", 
     ylab = "Frequency",
     col = "blue",
     border = "black")
```




```{r }
summary(probit_model)
```









