---
title: "Applied Metrics Empirical P-Set 2, Fall 2023"
output: pdf_document
date: "2023-11-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

### Load data

```{r }
# Clear all
rm(list = ls())

# load and require packages
pacman::p_load(foreign, tidyverse, here, randomForest, boot)

# Load data
data <- read.csv(here("data/andres_cleaned.csv"))
data$X <- NULL

```

### Question 2a - Relate structural equation and probit regression

Suppose we specify the potential outcome equations as $Y_i(1) = X'_i \theta_1 + \epsilon_{1i}$ and $Y_i(0) = X'_i \theta_0 + \epsilon_{0i}$, and the structural selection equation as $D_i = 1 \{u(X_i,Z_i) \geq V_i \}$. where $u(X_i,Z_i) - V_i$ is the latent utility derived from additional children.

The structural selection equation describes the switch from $D = 0$ to $D=1$ in terms of utility, i.e., whether, as a function of your observables and instrument, the family has a higher utility than some level of having more than three children. The propensity score quantifies this probability, capturing the likelihood, given your observables and instrument value, of having $D = 1$.

### Question 2b - Assumptions for MTE

Clarify a set of assumptions that the marginal treatment effect $M(x,u)$, $u \in (0,1)$ can be identified by the derivative of the OLS regression equation WRT the propensity score: $$MTE(x, u) = \frac{\partial}{\partial p}\bigg|_{p=u} \left( x'\beta_0 + px'\beta_1 + \kappa_1p + \kappa_2p^2 + \kappa_3p^3 \right)$$

The assumptions are: XXXXX

### Question 2c - Estimate propensity score

Estimate the propensity score as instructed above, and assess whether the coefficient of the instrumental variable in the probit regression is significantly different from zero or not.

Yes, we see that the z-value for the instrument is extremely large at 37, and very significant. So we have confidence that the instrument is predicting treatment.

```{r }
# Fit probit
probit_model <- glm(d ~ blackm + hispm + othracem + agem1 + agefstm + boy1st + z, 
                    data = data, 
                    family = binomial(link = "probit"))

# Predict propensity scores
data$prop_score <- predict(probit_model, type = "response")

summary(probit_model)
```

### Question 2d - Plot histogram of estimated propensity scores

```{r }
hist(data$prop_score, 
     main = "Histogram of Estimated Propensity Scores", 
     xlab = "Propensity Score", 
     ylab = "Frequency",
     col = "blue",
     border = "black")
```

```{r }
summary(probit_model)
```

### Question 2e - Estimate by OLS, bootstrap standard errors

Report the OLS estimates of the coefficients in the regression equation of (2) and their bootstrap standard errors with 1000 bootstrap replications.

```{r }
# Fit OLS
ols <- lm(y ~ blackm + hispm + othracem + agem1 + agefstm + boy1st + 
            prop_score*(blackm + hispm + othracem + agem1 + agefstm + boy1st) + 
            prop_score + (prop_score)^2 + prop_score^3, 
                    data = data)

summary(ols)

# Function to fit model on bootstrap sample
boot_fn <- function(data, index) {
  fit <- lm(y ~ blackm + hispm + othracem + agem1 + agefstm + boy1st + 
            prop_score*(blackm + hispm + othracem + agem1 + agefstm + boy1st) + 
            prop_score + I(prop_score^2) + I(prop_score^3), 
            data = data, subset = index)
  return(coef(fit))
}

# Perform bootstrap
set.seed(123) # For reproducibility
boot_results <- boot(data, boot_fn, R = 1000)

# Calculate standard errors
boot_se <- apply(boot_results$t, 2, sd)
```

### Question 2f - Calculate MTE and their CI

Plot the MTE estimates (as a function of u ∈ [0,1]) and pointwise confidence intervals,
holding X constant at its sample mean values. The confidence intervals are constructed
using the boostrap results.

```{r}
# Step 1: Calculate the means of the covariates
covariate_means <- colMeans(data[, c("blackm", "hispm", "othracem", "agem1", "agefstm", "boy1st")], na.rm = TRUE)

# Step 2: Fit the regression model (already done)

# Step 3 & 4: Use the fitted model to make predictions and calculate partial derivatives
prop_scores <- seq(0, 1, length.out = 100) # Creating a sequence from 0 to 1
marginal_effects <- numeric(length(prop_scores))

for (i in 1:length(prop_scores)) {
    # Create a new data frame with the mean values and the current prop_score
    new_data <- as.data.frame(t(covariate_means))
    new_data$prop_score <- prop_scores[i]
    new_data$I_prop_score_2 <- prop_scores[i]^2
    new_data$I_prop_score_3 <- prop_scores[i]^3
    
    # Predict using the new data
    pred <- predict(ols, new_data)

    # Calculate the marginal effect (partial derivative)
    # Assuming a linear relationship, this is equivalent to the coefficient of prop_score
    marginal_effects[i] <- coef(ols)["prop_score"] + 2 * coef(ols)["I(prop_score^2)"] * prop_scores[i] + 3 * coef(ols)["I(prop_score^3)"] * prop_scores[i]^2
}

# Step 5: Plot the results
plot(prop_scores, marginal_effects, type = "l", xlab = "Propensity Score", ylab = "Marginal Treatment Effect")

```

### Question 2x -

```{r}

```
